\documentclass[11pt]{llncs}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{epic,gastex}

\newcommand{\sa}{synchronizing automata}
\newcommand{\san}{synchronizing automaton}
\newcommand{\ssw}{a reset word of minimum length}


\DeclareMathOperator{\weight}{wg}

\DeclareSymbolFont{rsfscript}{OMS}{rsfs}{m}{n}
\DeclareSymbolFontAlphabet{\mathrsfs}{rsfscript}

\newcommand{\theoremtext}[1]{
For each $n$, there exists strongly connected \san\ with $n$ states,
such that it's shortest reset word is of length $#1$.
}

\newcommand{\lemmatext}[1]{
The word $#1$ is a reset word for the automaton $\mathrsfs{B}_n$.}


\begin{document}
\title{Slowly synchronizing automata and digraphs\thanks{Supported
by the Russian Foundation for Basic Research, grants 09-01-12142
and 10-01-00524, and by the Federal Education Agency of Russia,
grant 2.1.1/3537.}}

\titlerunning{Slowly synchronizing automata and digraphs}

\author{D. S. Ananichev \and V. V. Gusev \and M. V. Volkov}

\authorrunning{D. S. Ananichev, V. V. Gusev, M. V. Volkov}

\tocauthor{D. S. Ananichev, V. V. Gusev, M. V. Volkov
(Ekaterinburg, Russia)}

\institute{Department of Mathematics and Mechanics,\\
Ural State University, 620083 Ekaterinburg, RUSSIA\\
\email{Dmitry.Ananichev@usu.ru, vl.gusev@gmail.com,
Mikhail.Volkov@usu.ru}}

\maketitle

\begin{abstract}
We present several infinite series of \sa\
whose shortest reset words are close to conjectured by \v{C}ern\'{y} upper bound $(n - 1)^2$.
%Connection with primitive digraphs with big exponent.
\end{abstract}


\section{Background and overview}

A \emph{complete deterministic finite automaton} (DFA) is a triple
$\mathrsfs{A}=\langle Q,\Sigma,\delta\rangle$, where $Q$ and
$\Sigma$ are finite sets called the \emph{state set} and the
\emph{input alphabet} respectively, and $\delta:Q\times\Sigma\to
Q$ is a totally defined function called the \emph{transition
function}. Let $\Sigma^*$ stand for the collection of all finite
words over the alphabet $\Sigma$, including the empty word.
The function $\delta$ extends to a function $Q\times\Sigma^*\to Q$
(still denoted by $\delta$) in the following natural way: for every
$q\in Q$ and $w\in\Sigma^*$, we set
$$\delta(q,w)=\begin{cases}
q &\text{ if $w$ is empty},\\
\delta(\delta(q,v),a) &\text{ if $w=va$ for some $v\in\Sigma^*$ and $a\in\Sigma$}.
\end{cases}$$
Thus, via $\delta$, every word $w\in\Sigma^*$ acts on the set $Q$.

A DFA $\mathrsfs{A}=\langle Q,\Sigma,\delta\rangle$ is called
\emph{synchronizing} if the action of some word $w\in\Sigma^*$
resets $\mathrsfs{A}$, that is, leaves the automaton in one
particular state no matter at which state in $Q$ it is applied:
$\delta(q,w)=\delta(q',w)$ for all $q,q'\in Q$. Any word $w$
with this property is said to be a \emph{reset word} for the DFA.

Synchronizing automata serve as transparent and natural models of
error-resistant systems in many applications (coding theory, robotics,
testing of reactive systems) and also reveal interesting connections
with symbolic dynamics and other parts of mathematics. For a brief
introduction to the theory of \sa\ we refer the reader to the recent
surveys~\cite{Sa05,Vo08}. Here we focus on the so-called \v{C}ern\'{y}
conjecture that constitutes a major open problem in this area.

In~1964 \v{C}ern\'{y}~\cite{Ce64} constructed for each $n>1$ a \san\
$\mathrsfs{C}_n$ with $n$ states whose shortest reset word has length
$(n-1)^2$. Soon after that \v{C}ern\'{y} conjectured that these automata
represent the worst possible case, that is, every \san\ with $n$ states
can be reset by a word of length $(n-1)^2$. This imply looking conjecture
resists researchers' efforts for more than 40 years. Even though the
conjecture has been confirmed for various restricted classes of \sa\ (cf.\
\cite{Ep90,Du98,Ka03,Tr07,Vo09}, to mention a few advances only), no upper
bound of magnitude $O(n^2)$ for the minimum length of reset words for
$n$-state \sa\ is known in general---the best upper bound achieved so
far is $\frac{n^3-n}6$, see~\cite{Pi83}.

One of the difficulties that one encounters when approaching the
\v{C}ern\'{y} conjecture is that there are only very few examples
of \emph{extreme} \sa, that is, $n$-state \sa\ whose shortest reset
words have lengths $(n-1)^2$. In fact, the \v{C}ern\'{y} series
$\mathrsfs{C}_n$, $n=2,3,\dotsc$, is the only known infinite series
of extreme \sa. Besides that, we know only a few isolated examples
of such automata: up to now their complete list (see~\cite{Vo08})
consists of 8 automata of which the largest has 6 states. Moreover,
even \emph{slowly} \sa, that is, \sa\ whose shortest reset words have
lengths close to the \v{C}ern\'{y} bound are very rare. This empirical
observation is supported also by probabilistic arguments. For instance,
Higgins~\cite{Hi88} has shown that the probability that a composition of $2n$
random self-maps of a set of size $n$ is a constant map tends to~1 as $n$
goes to infinity. In terms of automata, Higgins's result means that
a random automaton with $n$ states and at least $2n$ input letters
has a reset word of length $2n$. For further results of the same flavor
see~\cite{SZ}. Thus, there is no hope to find new examples of slowly
\sa, to say nothing of a counterexample to the \v{C}ern\'{y}
conjecture, via a random sampling experiment.

We therefore have designed and performed a series of exhaustive search
experiments. Our experiments are briefly described in Section~\ref{experiments}
while the main body of the paper is devoted to a theoretical analysis
of their outcome. We concentrate on two principal issues. In Section~\ref{matrices}
we discuss a remarkable analogy between the distribution of lengths of shortest
reset words for \sa\ and the distribution of exponents of primitive non-negative
matrices. Section~\ref{sss} presents several new series of slowly \sa.
The ``initial'' examples in these series have been discovered in the course
of the experiments; then each example has been expanded to a series of automata
that have been proved to be slowly synchronizing. In our opinion, the proof
technique is also of interest; in fact, we provide a transparent and uniform
approach to all slowly \sa\ with 2 input letters, both new ones and already
known ones.



\section{Correlation between digraphs and automata}
\label{matrices}

\begin{theorem}\label{dulmage}
Apart from isomorphism, there is exactly one primitive graph $D$
on $n$ vertices for which $\gamma(D) = (n - 1)^2 + 1$, and exactly one
for which $\gamma(D) = (n - 1)^2$.
\end{theorem}

8 states:\\

\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}
49 & 48 & 47 & 46 & 45 & 44 & 43 & 42 & 41 & 40 \\
\hline
1  & 0  & 0  & 0  & 0  & 1  & 1  & 3  & 1  & 5  \\
\end{tabular}
\vspace{0.7cm}

9 states:\\
\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|c|c|c|c|}
64 & 63 & 62 & 61 & 60 & 59 & 58 & 57 & 56 & 55 & 54 & 53 & 52 & 51 \\
\hline
1  & 0  & 0  & 0  & 0  & 0  & 1  & 2  & 3  & 0  & 0  & 0  & 4  & 4   \\
\end{tabular}


\section{Main results and slowly sinchronizing series}
\label{sss}

\begin{theorem}
For each $n$, there exists strongly connected \san\ with $n$ states,
such that it's shortest reset word is of length $n^2-3n+3$.
\end{theorem}

\begin{lemma}
\label{lemma:anan}
The word $(ab^{n - 2})^{n - 2}a$ is a reset word for the automaton $\mathrsfs{B}_n$.
\end{lemma}

\begin{figure}[ht]
\begin{center}
\unitlength .7mm
\begin{picture}(112,76)(0,-86)
\gasset{Nadjustdist=1.5} \node(n0)(56.0,-16.0){1}
\node(n1)(24.0,-40.0){0} \node(n2)(88.0,-40.0){2}
\node(n3)(36.0,-72.0){4} \node(n4)(76.0,-72.0){3}
\drawedge[ELdist=2.0](n1,n0){$b$} \drawedge[ELdist=1.5](n2,n4){$a, b$}
\drawedge[ELdist=1.7](n0,n2){$a, b$} \drawedge[ELdist=2.0](n4,n3){$a, b$}
\drawedge[ELdist=1.7](n3,n1){$a, b$} \drawedge[ELdist=2.0](n1,n2){$a$}
\end{picture}
\end{center}
\caption{The automaton $\mathrsfs{B}_5$}\label{fig:anan}
\end{figure}

The states of the automaton $\mathrsfs{B}_n$
are the residues modulo $n$ and its input letters $a$ and $b$ act
as follows:
$$
 \delta(m,a)=
 \begin{cases}
  m + 2\!\!\pmod{n} & \text{for $m = 0$}, \\
  m+1\!\!\pmod{n} & \text{for $m \neq 0$};
  \end{cases}
\qquad \delta(m,b)=m+1\!\!\pmod{n}.
$$
The smallest automaton in the series is shown in Fig.~\ref{fig:anan}.

\begin{proof}
Let $w$ be \ssw . It is clear that action of $w$ leaves automaton $\mathrsfs{B}_n$ in state 2.
For any $u \in \Sigma^*$ word $uw$ is synchronizing too and it also brings automaton to the state 2.
In other words, there is a path in underlying
digraph from any vertex to vertex 2 of length $p$, if $p \geq |w|$.
Let us consider paths from 2 to 2. Every one of them is the combination of cycles of length $n$ and $n - 1$.
Thus for any $p \geq |w|$ the following is true: $p = xn + y(n - 1)$, where $x$ and $y$ are nonnegative integers.
It is well known(See smth) that $n(n - 1) - n - (n - 1) = n^2 - 3n + 1$ is the greatest number that could not be represented
as a nonegative integer combination of $n$ and $n - 1$. Thus $|w| \geq n^2 - 3n + 2$.
Suppose that $|w| = n^2 - 3n + 2$.
%and  then there is a path of such length from state $1$ to state $2$.
If we read this word from state $1$ then we get a path of length $n^2 - 3n + 1$ from $2$ to $2$ which is not possible due to previous
statement. Thus $|w| \geq n^2 - 3n + 3$ but this bound is tight by lemma~\ref{lemma:anan}.
\end{proof}
\newpage

\begin{theorem}\label{theo}
\theoremtext{n^2-3n+3}
\end{theorem}

\begin{lemma}
\lemmatext{(ab^{n - 2})^{n - 2}a}
\end{lemma}

\begin{figure}[ht]
\begin{center}
\unitlength .7mm
\begin{picture}(112,76)(0,-86)
\gasset{Nadjustdist=1.5} \node(n0)(56.0,-16.0){0}
\node(n1)(24.0,-40.0){4} \node(n2)(88.0,-40.0){1}
\node(n3)(36.0,-72.0){3} \node(n4)(76.0,-72.0){2}
\drawloop[ELdist=1.5,loopangle=33.34](n2){$a$}
\drawloop[ELdist=2.4,loopangle=320.0](n4){$a$}
\drawedge[ELdist=2.0](n1,n0){$b$} \drawedge[ELdist=1.5](n2,n4){$b$}
\drawedge[ELdist=1.7](n0,n2){$a, b$} \drawedge[ELdist=2.0](n4,n3){$b$}
\drawedge[ELdist=1.7](n3,n1){$b$}
%\drawedge[ELdist=2.0](n1,n2){$a$}
%\drawedge[ELdist=2.0](n0,n4){$a$}
\drawloop[ELdist=1.5,loopangle=144.55](n1){$a$}
\drawloop[ELdist=1.5,loopangle=226.55](n3){$a$}
\end{picture}
\end{center}
\caption{The automaton $\mathrsfs{B}_5$}\label{B5}
\end{figure}

The states of the automaton $\mathrsfs{B}_n$
are the residues modulo $n$ and its input letters $a$ and $b$ act
as follows:
$$
 \delta(m,a)=
 \begin{cases}
  m + 2 \!\!\pmod{n} & \text{for $m = 0$}, \\
  m \!\!\pmod{n} & \text{for $m \neq 0$};
  \end{cases}
\qquad \delta(m,b)=m+1\!\!\pmod{n}.
$$
The smallest automaton in the series is shown in Fig.~\ref{B5}.


\newpage







\begin{lemma}
\lemmatext{(ab^{n - 2})^{n - 2}ba}
\end{lemma}

\begin{theorem}\label{theo}
\theoremtext{n^2-3n+4}
\end{theorem}


\begin{figure}[ht]
\begin{center}
\unitlength .7mm
\begin{picture}(112,76)(0,-86)
\gasset{Nadjustdist=1.5} \node(n0)(56.0,-16.0){1}
\node(n1)(24.0,-40.0){0} \node(n2)(88.0,-40.0){2}
\node(n3)(36.0,-72.0){4} \node(n4)(76.0,-72.0){3}
\drawedge[ELdist=2.0](n1,n0){$b$} \drawedge[ELdist=1.5](n2,n4){$a, b$}
\drawedge[ELdist=1.7](n0,n2){$b$} \drawedge[ELdist=2.0](n4,n3){$a, b$}
\drawedge[ELdist=1.7](n3,n1){$a, b$} \drawedge[ELpos=40, ELdist=2.0](n1,n2){$a$}
\drawedge[ELpos=60,ELdist=2.0](n0,n4){$a$}
\end{picture}
\end{center}
\caption{The automaton $\mathrsfs{B}_5$}\label{B5}
\end{figure}

The states of the automaton $\mathrsfs{B}_n$
are the residues modulo $n$ and its input letters $a$ and $b$ act
as follows:
$$
 \delta(m,a)=
 \begin{cases}
  m + 2\!\!\pmod{n} & \text{for $m = 0,1$}, \\
  m+1\!\!\pmod{n} & \text{for $1< m<n$};
  \end{cases}
\qquad \delta(m,b)=m+1\!\!\pmod{n}.
$$
The smallest automaton in the series is shown in Fig.~\ref{B5}.

\newpage

\begin{theorem}
\theoremtext{n^2-3n+2}
\end{theorem}

\begin{lemma}
\lemmatext{(ba^{n - 1})^{n - 3}ba}
\end{lemma}

\begin{figure}[ht]
\begin{center}
\unitlength .7mm
\begin{picture}(112,76)(0,-86)
\gasset{Nadjustdist=1.5} \node(n0)(56.0,-16.0){1}
\node(n1)(24.0,-40.0){0} \node(n2)(88.0,-40.0){2}
\node(n3)(36.0,-72.0){4} \node(n4)(76.0,-72.0){3}
\drawedge[ELdist=2.0](n1,n0){$b$} \drawedge[ELdist=1.5](n2,n4){$a, b$}
\drawedge[ELdist=1.7](n0,n2){$a$} \drawedge[ELdist=2.0](n4,n3){$a, b$}
\drawedge[ELdist=1.7](n3,n1){$a, b$} \drawedge[ELpos=40, ELdist=2.0](n1,n2){$a$}
\drawedge[ELpos=60,ELdist=2.0](n0,n4){$b$}
\end{picture}
\end{center}
\caption{The automaton $\mathrsfs{B}_5$}\label{B5}
\end{figure}

The states of the automaton $\mathrsfs{B}_n$
are the residues modulo $n$ and its input letters $a$ and $b$ act
as follows:
$$
 \delta(m,a)=
 \begin{cases}
  m+2\!\!\pmod{n} & \text{for $m = 1$}, \\
  m+1\!\!\pmod{n} & \text{for $m \neq 0$};
  \end{cases}
\qquad \delta(m,b)=
  \begin{cases}
  m + 2\!\!\pmod{n} & \text{for $m = 1$}, \\
  m+1\!\!\pmod{n} & \text{for $m \neq 1$};
  \end{cases}.
$$
The smallest automaton in the series is shown in Fig.~\ref{B5}.

\newpage


\begin{theorem}\label{theo}
\theoremtext{n^2-3n+3}
\end{theorem}

\begin{lemma}
\lemmatext{(ab^{n - 2})^{n - 2}a}
\end{lemma}

\begin{figure}[ht]
\begin{center}
\unitlength .7mm
\begin{picture}(112,76)(0,-86)
\gasset{Nadjustdist=1.5} \node(n0)(56.0,-16.0){0}
\node(n1)(24.0,-40.0){4} \node(n2)(88.0,-40.0){1}
\node(n3)(36.0,-72.0){3} \node(n4)(76.0,-72.0){2}
\drawloop[ELdist=1.5,loopangle=33.34](n2){$a$}
\drawloop[ELdist=2.4,loopangle=320.0](n4){$a$}
\drawedge[ELdist=2.0](n1,n0){$b$} \drawedge[ELdist=1.5](n2,n4){$b$}
\drawedge[ELdist=1.7](n0,n2){$b$} \drawedge[ELdist=2.0](n4,n3){$b$}
\drawedge[ELdist=1.7](n3,n1){$b$}
%\drawedge[ELdist=2.0](n1,n2){$a$}
\drawedge[ELdist=2.0](n0,n4){$a$}
\drawloop[ELdist=1.5,loopangle=144.55](n1){$a$}
\drawloop[ELdist=1.5,loopangle=226.55](n3){$a$}
\end{picture}
\end{center}
\caption{The automaton $\mathrsfs{B}_5$}\label{B5}
\end{figure}

The states of the automaton $\mathrsfs{B}_n$
are the residues modulo $n$ and its input letters $a$ and $b$ act
as follows:
$$
 \delta(m,a)=
 \begin{cases}
  m + 2 \!\!\pmod{n} & \text{for $m = 0$}, \\
  m \!\!\pmod{n} & \text{for $m \neq 0$};
  \end{cases}
\qquad \delta(m,b)=m+1\!\!\pmod{n}.
$$
The smallest automaton in the series is shown in Fig.~\ref{B5}.

\newpage


\begin{theorem}\label{theo}
\theoremtext{n^2-3n+2}
\end{theorem}

\begin{lemma}
\lemmatext{(aab^{n - 2})^{n - 3}aa}
\end{lemma}

\begin{figure}[ht]
\begin{center}
\unitlength .7mm
\begin{picture}(112,76)(0,-86)
\gasset{Nadjustdist=1.5} \node(n0)(56.0,-16.0){1}
\node(n1)(24.0,-40.0){0} \node(n2)(88.0,-40.0){2}
\node(n3)(36.0,-72.0){4} \node(n4)(76.0,-72.0){3}
\drawloop[ELdist=1.5,loopangle=33.34](n2){$a$}
\drawloop[ELdist=2.4,loopangle=320.0](n4){$a$}
\drawedge[ELdist=2.0](n1,n0){$a$} \drawedge[ELdist=1.5](n2,n4){$b$}
\drawedge[ELdist=1.7](n0,n2){$a, b$} \drawedge[ELdist=2.0](n4,n3){$b$}
\drawedge[ELdist=1.7](n3,n1){$b$}
\drawedge[ELdist=2.0](n1,n2){$b$}
%\drawedge[ELdist=2.0](n0,n4){$a$}
%\drawloop[ELdist=1.5,loopangle=144.55](n1){$a$}
\drawloop[ELdist=1.5,loopangle=226.55](n3){$a$}
\end{picture}
\end{center}
\caption{The automaton $\mathrsfs{B}_5$}\label{B5}
\end{figure}

The states of the automaton $\mathrsfs{B}_n$
are the residues modulo $n$ and its input letters $a$ and $b$ act
as follows:
$$
 \delta(m,a)=
 \begin{cases}
  m + 1 \!\!\pmod{n} & \text{for $m = 0, 1$}, \\
  m \!\!\pmod{n} & \text{for $1 < m < n$};
  \end{cases}
\qquad
 \delta(m,b)=\begin{cases}
  m + 2 \!\!\pmod{n} & \text{for $m = 0$}, \\
  m + 1 \!\!\pmod{n} & \text{for $m \neq 0$};
  \end{cases}.
$$
The smallest automaton in the series is shown in Fig.~\ref{B5}.

\newpage

\begin{theorem}\label{theo}
\theoremtext{n^2-4n+6}
\end{theorem}

\begin{lemma}
\lemmatext{b(ab^{n - 2})^{n - 3}ab}
\end{lemma}


\begin{figure}[th]
\unitlength=.7mm
\begin{center}
\begin{picture}(100,70)(0,-80)
\gasset{Nadjustdist=1.5} \node(n0)(56.0,-16.0){0}
\node(n1)(24.0,-40.0){4} \node(n2)(88.0,-40.0){1}
\node(n3)(36.0,-72.0){3} \node(n4)(76.0,-72.0){2}
\drawloop[ELdist=2.4,loopangle=320.0](n4){$a$}
\drawedge[ELdist=2.0](n1,n0){$b$} \drawedge[ELdist=1.5](n2,n4){$b$}
\drawedge[ELdist=1.7](n0,n2){$b$} \drawedge[ELdist=2.0](n4,n3){$b$}
\drawedge[ELdist=1.7](n3,n1){$b$}
\node[NLangle=0.0](n16)(56.0,-48.0){5}
%\drawedge[ELdist=1.6](n1,n16){$a$}
\drawedge[ELdist=1.7](n16,n4){$b$}

\drawedge[curvedepth=2](n0,n16){$a$}
\drawedge[curvedepth=2](n16,n0){$a$}

%\drawloop[ELdist=1.5,loopangle=90](n0){$a$}
%\drawloop[ELdist=1.5](n16){$b$}
%\drawedge[ELdist=1.5,curvedepth=33](n3,n0){$a$}
\drawloop[ELdist=1.5,loopangle=144.55](n1){$a$}
\drawloop[ELdist=1.5,loopangle=226.55](n3){$a$}
\drawloop[ELdist=1.5,loopangle=33.34](n2){$a$}
\end{picture}
\end{center}
\caption{The automaton $\mathrsfs{B}^{\star}_6$}\label{B-star-6}
\end{figure}

\newpage

\begin{theorem}\label{theo}
\theoremtext{n^2-4n+6}
\end{theorem}

\begin{lemma}
\lemmatext{b(ab^{n - 2})^{n - 3}ba}
\end{lemma}


\begin{figure}[th]
\unitlength=.7mm
\begin{center}
\begin{picture}(100,70)(0,-80)
\gasset{Nadjustdist=1.5} \node(n0)(56.0,-16.0){0}
\node(n1)(24.0,-40.0){4} \node(n2)(88.0,-40.0){1}
\node(n3)(36.0,-72.0){3} \node(n4)(76.0,-72.0){2}
\drawloop[ELdist=2.4,loopangle=320.0](n4){$a$}
\drawedge[ELdist=2.0](n1,n0){$b$} \drawedge[ELdist=1.5](n2,n4){$a, b$}
\drawedge[ELdist=1.7](n0,n2){$b$} \drawedge[ELdist=2.0](n4,n3){$b$}
\drawedge[ELdist=1.7](n3,n1){$b$}
\node[NLangle=0.0](n16)(56.0,-48.0){5}
\drawedge[ELdist=1.7](n16,n4){$b$}

\drawedge[curvedepth=2](n0,n16){$a$}
\drawedge[curvedepth=2](n16,n0){$a$}

\drawloop[ELdist=1.5,loopangle=144.55](n1){$a$}
\drawloop[ELdist=1.5,loopangle=226.55](n3){$a$}
\end{picture}
\end{center}
\caption{The automaton $\mathrsfs{B}^{\star}_6$}\label{B-star-6}
\end{figure}

\section{Experiments}
\label{experiments}

\begin{thebibliography}{99}
\bibitem{AVZ}
Dimitry S. Ananichev, Mikhail V. Volkov, Yu. I. Zaks,
Synchronizing automata with a letter of deficiency 2. Theor.\ Comput.\ Sci.
376 (2007)  30--41.

\bibitem{Ce64}
\v{C}ern\'{y}, J.: Pozn\'{a}mka k homog\'{e}nnym eksperimentom s
kone\v{c}n\'{y}mi automatami. Matematicko-fyzikalny \v{C}asopis
Slovensk.\ Akad.\ Vied 14(3) 208--216 (1964) (in Slovak)

\bibitem{Du98}
Dubuc, L.: Sur les automates circulaires et la conjecture de
\v{C}ern\'y. RAIRO Inform.\ Th\'eor.\ Appl. 32, 21--34 (1998) (in
French)

\bibitem{Ep90}
Eppstein, D.: Reset sequences for monotonic automata. SIAM J.
Comput. 19, 500--510 (1990)

\bibitem{Hi88}
Higgins, P.M.: The range order of a product of $i$ transformations
from a finite full transformation semigroup, Semigroup Forum 37, 31--36
(1988)

\bibitem{Ka03}
Kari, J.: Synchronizing finite automata on Eulerian digraphs.
Theoret.\ Comput.\ Sci. 295, 223--232 (2003)

\bibitem{Pi83}
Pin, J.-E.: On two combinatorial problems arising from automata
theory. Ann.\ Discrete Math. 17, 535--548 (1983)

\bibitem{Sa03}
A.\,Salomaa, Composition sequences for functions over a finite
domain, Theoret.\ Comput.\ Sci. 292 (2003) 263--281.

\bibitem{Sa05}
Sandberg, S.: Homing and synchronizing sequences. In: Broy, M.
et~al. (eds.), Model-Based Testing of Reactive Systems. Lect.\
Notes Comput.\ Sci., vol.\,3472, pp.\,5--33. Springer, Heidelberg
(2005)

\bibitem{SZ}
Skvortsov E., Zaks Yu.: Synchronizing random automata. Submitted;
proceedings version in: Rigo, M. (ed.), AutoMathA 2009, Universit\'e
de Li\`ege (2009)

\bibitem{Tr06}
Trahtman, A.: An efficient algorithm finds noticeable trends
and examples concerning the \v{C}ern\'y conjecture. In:
Kr\'alovi\v{c}, R.; Urzyczyn, P. (eds.), Mathematical Foundations
of Computer Science. Lect.\ Notes Comput.\ Sci., vol.\,4162, pp.\,789--800
Springer, Heidelberg (2006)

\bibitem{Tr07}
Trahtman, A.: The \v{C}ern\'y conjecture for aperiodic automata.
Discrete Math.\ Theor.\ Comput.\ Sci. 9(2), 3--10 (2007)

\bibitem{Vo08}
Volkov, M.V.: Synchronizing automata and the \v{C}ern\'{y}
conjecture. In: Mart\'\i{}n-Vide, C.; Otto, F.; Fernau, H. (eds.),
Languages and Automata: Theory and Applications. Lect.\ Notes
Comput.\ Sci., vol.\,5196, pp.\,11--27.  Springer, Heidelberg (2008)

\bibitem{Vo09}
Volkov, M.V.: Synchronizing automata preserving a chain of partial
orders. Theoret.\ Comput.\ Sci. 410, 2992--2998 (2009)
\end{thebibliography}


\section*{Appendix}

\end{document}
